---
title: "cfDNA vs gDNA 5-hmC at Screening as Predictors of MRD by NGS at Cycle 8"
subtitle: "Predictive-modeling complement to the Bayesian DP-mixture comparison"
author: "Saurabh Bhandari, Yuan Ji, Brian C.-H. Chiu, and others"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage{array}
  - \usepackage[utf8]{inputenc}
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.pos   = "H",
  fig.align = "center",
  out.width = "85%"
)

set.seed(20260609)

suppressPackageStartupMessages({
  # Load Bioconductor packages FIRST, so their exports (rename, select, ...)
  # are masked by the dplyr/tidyr versions loaded later. Otherwise, dplyr::
  # qualifiers are needed everywhere.
  library(DESeq2)
  library(SummarizedExperiment)
  library(dplyr)
  library(tidyr)
  library(readxl)
  library(mclust)
  library(glmnet)
  library(pROC)
  library(ggplot2)
  library(knitr)
  library(kableExtra)
})

base_dir  <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
                           mustWork = FALSE)
data_dir  <- file.path(base_dir, "KRd trial", "5hmC data")
clin_path <- file.path(base_dir, "KRd trial", "Clinical data",
                       "KRd 12_1725 data_2_14_2023 from Ben without PHI.xlsx")

fig_out <- file.path(base_dir, "nature_manuscript", "figures")
dir.create(fig_out, showWarnings = FALSE, recursive = TRUE)

# Helper: save every figure to nature_manuscript/figures/ as both PDF and PNG.
# Use base R devices via `device = grDevices::pdf` / `grDevices::png` rather
# than ragg, which can hit "Graphics API version mismatch" inside knitr.
save_both <- function(plot, basename, width = 7, height = 5, dpi = 300) {
  pdf_path <- file.path(fig_out, paste0(basename, ".pdf"))
  png_path <- file.path(fig_out, paste0(basename, ".png"))
  ggsave(pdf_path, plot, width = width, height = height, device = grDevices::pdf)
  ggsave(png_path, plot, width = width, height = height,
         device = grDevices::png, dpi = dpi, units = "in")
  invisible(list(pdf = pdf_path, png = png_path))
}
```

# Section 1. cfDNA vs gDNA at Screening as predictors of MRD by NGS at Cycle 8

## Introduction

The Bayesian Dirichlet-process-mixture comparison in the main text reaches a
verdict: under the source-DP $\mathcal{M}_0$ null, the
data favour hierarchical coupling of cfDNA and gDNA mixture weights ($\Delta$WAIC
$\approx +3.3 \times 10^{6}$). This document provides a *complementary,
predictive* check at the patient level: **can cfDNA 5-hmC measured at Screening
predict MRD-by-NGS status at Cycle 8 about as well as gDNA 5-hmC measured at
Screening?** If a cfDNA-only model performs comparably to a gDNA-only model on
this prediction task, that is direct clinical-utility support for cfDNA as a
non-invasive early-monitoring biomarker.

We fit three penalised logistic models on the 17 patients with paired
Screening cfDNA + gDNA 5-hmC profiles AND a recorded MRD-NGS C8 outcome:

  (i)   **cfDNA-only**: predictors = age + screened cfDNA 5-hmC genes;
  (ii)  **gDNA-only**: predictors = age + screened gDNA 5-hmC genes;
  (iii) **cfDNA + gDNA combined**: predictors = age + screened cfDNA genes +
        screened gDNA genes.

Age is included unpenalised.
We use leave-one-out cross-validation (LOO-CV) with the two-stage gene screen
(univariate logistic Wald-p filter + GMM on $|\hat\beta|$) executed **inside**
each LOO fold so that no held-out information leaks into screening.

This analysis is framed as **exploratory and hypothesis-generating**, not as
confirmatory predictive validation: with $n=17$ and a 7-vs-10 class split,
DeLong 95% confidence intervals on AUROCs are inherently wide (~0.30), and a
non-significant cfDNA-vs-gDNA difference is consistent with both genuine
equivalence and an underpowered comparison.

\newpage

## Data ingestion and cohort construction

```{r load-data}
cf_counts <- readRDS(file.path(data_dir, "kRd-cfDNA_genebody_count.RDS"))
g_counts  <- readRDS(file.path(data_dir, "kRd-gDNA_genebody_count.RDS"))
cf_key    <- read.csv(file.path(data_dir, "kRd-cfDNA_sample_key.csv"),
                      stringsAsFactors = FALSE)
g_key     <- read.csv(file.path(data_dir, "kRd-gDNA_sample_key.csv"),
                      stringsAsFactors = FALSE)
clin      <- suppressWarnings(
  read_excel(clin_path, sheet = "12_1725 cfDNA Dataset")
)
```

```{r build-cohort}
# Restrict each key to Screening
cf_scr <- dplyr::filter(cf_key, Timepoint == "Screening")
g_scr  <- dplyr::filter(g_key,  Timepoint == "Screening")

# Paired patients = Study.ID present in BOTH Screening keys
paired_ids <- intersect(cf_scr$Study.ID, g_scr$Study.ID)

# Clinical: MRD-C8 + Age, keyed on Study.ID
clin_min <- clin %>%
  dplyr::transmute(
    Study.ID  = `Study number`,
    mrd_neg   = `MRD post 8 cycles by NGS (Adaptive) negative =1`,
    age       = Age
  ) %>%
  dplyr::filter(!is.na(mrd_neg))

# Cohort = paired patients with MRD-C8 + Age
cohort <- data.frame(Study.ID = paired_ids, stringsAsFactors = FALSE) %>%
  dplyr::inner_join(clin_min, by = "Study.ID") %>%
  dplyr::arrange(Study.ID)

stopifnot(nrow(cohort) == 17)

# Outcome coding: y = 1 if MRD-POSITIVE (residual disease, "worse" outcome).
# clin's "MRD ... negative = 1" so y = 1 - mrd_neg.
cohort$y01 <- 1L - as.integer(cohort$mrd_neg)
cohort$y   <- factor(cohort$y01, levels = c(0L, 1L),
                     labels = c("MRDneg", "MRDpos"))

# Per-patient Screening sample barcodes for the two count matrices.
# Note: explicit dplyr:: qualifiers because S4Vectors (loaded by DESeq2)
# also exports `rename` and `select`, which mask the dplyr versions.
cf_lookup <- dplyr::select(cf_scr, Study.ID, Assigned.ID)
g_lookup  <- dplyr::select(g_scr,  Study.ID, Assigned.ID)

cohort <- cohort %>%
  dplyr::left_join(dplyr::rename(cf_lookup, cf_barcode = Assigned.ID),
                   by = "Study.ID") %>%
  dplyr::left_join(dplyr::rename(g_lookup,  g_barcode  = Assigned.ID),
                   by = "Study.ID")

# Subset the count matrices to the 17 paired Screening samples,
# in the canonical patient order defined by `cohort`.
cf_counts_cohort <- cf_counts[, cohort$cf_barcode, drop = FALSE]
g_counts_cohort  <- g_counts[,  cohort$g_barcode,  drop = FALSE]
stopifnot(ncol(cf_counts_cohort) == 17, ncol(g_counts_cohort) == 17)
colnames(cf_counts_cohort) <- cohort$Study.ID
colnames(g_counts_cohort)  <- cohort$Study.ID
```

```{r table1-cohort}
tab1_summary <- data.frame(
  Characteristic = c(
    "n (paired Screening cfDNA + gDNA with MRD-C8)",
    "Age (years), mean (SD)",
    "Age (years), range",
    "MRD-positive at C8 (n, %)",
    "MRD-negative at C8 (n, %)"
  ),
  Value = c(
    sprintf("%d", nrow(cohort)),
    sprintf("%.1f (%.1f)", mean(cohort$age), sd(cohort$age)),
    sprintf("%d - %d", min(cohort$age), max(cohort$age)),
    sprintf("%d (%.1f%%)",
            sum(cohort$y01 == 1),
            100 * mean(cohort$y01 == 1)),
    sprintf("%d (%.1f%%)",
            sum(cohort$y01 == 0),
            100 * mean(cohort$y01 == 0))
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  tab1_summary, format = "latex", booktabs = TRUE, escape = TRUE,
  caption = "Cohort characteristics. Outcome is MRD-NGS positivity at Cycle 8."
) %>% kable_styling(latex_options = c("HOLD_position"))
```

\newpage

## Per-source filtering and variance-stabilising transformation

VST is applied **separately** to cfDNA and gDNA so that the
predictive performance of each source reflects only its own variance structure.
Filtering criterion (mirrors `R/02_preprocess.R`): remove genes with fewer
than 10 counts in more than 5% of cohort samples within each source.

```{r filter-and-vst}
filter_counts <- function(M, min_count = 10, frac_max_low = 0.05) {
  n_samp <- ncol(M)
  thr <- ceiling(n_samp * frac_max_low)
  low_per_gene <- rowSums(M < min_count)
  keep <- low_per_gene <= thr
  M[keep, , drop = FALSE]
}

cf_counts_f <- filter_counts(cf_counts_cohort)
g_counts_f  <- filter_counts(g_counts_cohort)

vst_per_source <- function(M, label) {
  meta <- data.frame(sample = colnames(M),
                     row.names = colnames(M),
                     stringsAsFactors = FALSE)
  dds <- DESeqDataSetFromMatrix(
    countData = round(M),
    colData   = meta,
    design    = ~ 1
  )
  vsd <- vst(dds, blind = TRUE)
  t(SummarizedExperiment::assay(vsd))   # samples x genes
}

X_cf <- vst_per_source(cf_counts_f, "cfDNA")
X_g  <- vst_per_source(g_counts_f,  "gDNA")

stopifnot(nrow(X_cf) == 17, nrow(X_g) == 17)
```

```{r vst-summary}
vst_summary <- data.frame(
  Source = c("cfDNA", "gDNA"),
  `Raw genes`            = c(nrow(cf_counts_cohort), nrow(g_counts_cohort)),
  `After count filter`   = c(ncol(X_cf), ncol(X_g)),
  `% retained`           = c(
    sprintf("%.1f%%", 100 * ncol(X_cf) / nrow(cf_counts_cohort)),
    sprintf("%.1f%%", 100 * ncol(X_g)  / nrow(g_counts_cohort))
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(
  vst_summary, format = "latex", booktabs = TRUE, escape = TRUE,
  caption = "Gene counts before and after per-source filtering (10-count threshold in >5% of 17 samples). DESeq2 VST applied per source with design = ~1, blind = TRUE."
) %>% kable_styling(latex_options = c("HOLD_position"))
```

\newpage

## Two-stage gene screening (executed inside each LOO fold)

Stage 1 fits a univariate logistic regression of MRD-C8 status on each gene
$j$: $\mathrm{logit}\,\Pr(y = 1) = \alpha_j + \beta_j x_{ij}$. Genes with Wald
$p$-value $<\alpha$ (default 0.05) survive, capped at the 200 most significant.

Stage 2 fits a 3-component Gaussian mixture (via `mclust::Mclust`) to the
absolute coefficients $|\hat\beta_j|$ of the Stage-1 survivors and keeps the
cluster with the largest mean $|\hat\beta|$ ("most predictive"). Defensive
fallbacks: with $\leq 3$ survivors, skip GMM and keep all; if `mclust` errors,
fall back to k-means with `centers = 3`.

The training-set means and SDs returned by the screen are used to standardize
both the training rows and the held-out row, preventing test-set leakage into
standardization.

```{r screen-fn}
# .logit_screen_one: univariate logit Wald p-value + slope for one gene.
# Returns c(beta, p). NA on non-convergence.
.logit_screen_one <- function(x, y01) {
  fit <- tryCatch(
    suppressWarnings(glm.fit(cbind(1, x), y01,
                             family = binomial())),
    error = function(e) NULL)
  if (is.null(fit) || !fit$converged) return(c(NA_real_, NA_real_))
  beta <- fit$coefficients[2]
  V <- tryCatch(summary.glm(structure(fit, class = c("glm","lm")))$cov.scaled,
                error = function(e) NULL)
  if (is.null(V) || nrow(V) < 2 || is.na(V[2,2]) || V[2,2] <= 0)
    return(c(beta, NA_real_))
  se <- sqrt(V[2,2]); z <- beta / se
  c(beta, 2 * stats::pnorm(-abs(z)))
}

# screen_genes(): returns the kept-gene names plus train-set scale info.
screen_genes <- function(Xtr, y01tr,
                         alpha = 0.05, cap = 200, G = 3, min_for_gmm = 4L) {
  stopifnot(nrow(Xtr) == length(y01tr))
  res <- apply(Xtr, 2, .logit_screen_one, y01 = y01tr)
  beta_all <- res[1, ]
  p_all    <- res[2, ]
  ord <- order(p_all, na.last = NA)
  if (length(ord) == 0L) {
    # nothing converged; keep most-variable gene as a safe fallback
    v   <- apply(Xtr, 2, sd)
    g0  <- names(v)[which.max(v)]
    keep_genes <- g0
  } else {
    survivors <- ord[which(p_all[ord] < alpha)]
    if (length(survivors) > cap) survivors <- survivors[seq_len(cap)]
    if (length(survivors) == 0L) survivors <- ord[1]
    surv_genes <- colnames(Xtr)[survivors]
    surv_absb  <- abs(beta_all[survivors])
    # Stage 2: GMM on |beta|, keep largest-mean cluster
    if (length(survivors) < min_for_gmm) {
      keep_genes <- surv_genes
    } else {
      cl <- tryCatch({
        gmm <- mclust::Mclust(surv_absb, G = G, verbose = FALSE)
        gmm$classification
      }, error = function(e) {
        # k-means fallback
        km <- tryCatch(stats::kmeans(matrix(surv_absb, ncol = 1),
                                     centers = G, nstart = 5),
                       error = function(e2) NULL)
        if (is.null(km)) return(rep(1L, length(surv_absb)))
        km$cluster
      })
      mu_by_cl <- tapply(surv_absb, cl, mean)
      top_cl   <- as.integer(names(mu_by_cl)[which.max(mu_by_cl)])
      keep_idx <- which(cl == top_cl)
      keep_genes <- surv_genes[keep_idx]
      if (length(keep_genes) == 0L) keep_genes <- surv_genes
    }
  }
  Xtr_keep <- Xtr[, keep_genes, drop = FALSE]
  mu_g     <- colMeans(Xtr_keep)
  sd_g     <- apply(Xtr_keep, 2, sd)
  sd_g[sd_g <= 1e-8] <- 1
  list(
    genes     = keep_genes,
    center    = mu_g,
    scale     = sd_g,
    beta_full = beta_all[keep_genes],
    p_full    = p_all[keep_genes],
    n_stage1  = sum(p_all < alpha, na.rm = TRUE),
    n_kept    = length(keep_genes)
  )
}
```

\newpage

## Nested LOO-CV prediction driver

For each held-out patient $i \in \{1, \ldots, 17\}$:

1. **Per-source screening** on the 16 training patients (independent for cfDNA
   and gDNA in the combined model, per the chosen design).
2. **Standardize** each kept gene using train mean and SD; standardize age the
   same way.
3. Fit `cv.glmnet` (elastic net, $\alpha = 0.5$, inner 5-fold CV,
   `type.measure = "deviance"`) with `penalty.factor = 0` for age and 1 for
   every gene.
4. Predict the held-out patient's MRD-positive probability at `s = "lambda.min"`.

```{r loo-driver}
loo_predict <- function(X_list, y01, age,
                        alpha_enet = 0.5, alpha_screen = 0.05, cap_screen = 200) {
  n <- length(y01)
  stopifnot(all(sapply(X_list, nrow) == n), length(age) == n)
  p_hat       <- numeric(n)
  n_kept_log  <- matrix(NA_real_, nrow = n, ncol = length(X_list),
                        dimnames = list(NULL, names(X_list)))
  fold_bal    <- matrix(NA_integer_, n, 2,
                        dimnames = list(NULL, c("MRDneg", "MRDpos")))
  screens_log <- vector("list", n)

  for (i in seq_len(n)) {
    tr <- setdiff(seq_len(n), i)
    y_tr <- y01[tr]

    fold_bal[i, ] <- c(sum(y_tr == 0), sum(y_tr == 1))

    # Per-source screening on training rows only
    screens <- lapply(X_list, function(X) {
      screen_genes(X[tr, , drop = FALSE], y_tr,
                   alpha = alpha_screen, cap = cap_screen, G = 3)
    })
    screens_log[[i]] <- lapply(screens, function(s) s$genes)
    n_kept_log[i, ] <- sapply(screens, function(s) length(s$genes))

    # Build training and held-out feature matrices
    age_mu <- mean(age[tr]); age_sd <- sd(age[tr])
    age_sd <- if (age_sd > 0) age_sd else 1
    age_tr <- (age[tr] - age_mu) / age_sd
    age_te <- (age[i]  - age_mu) / age_sd

    gene_block_tr <- do.call(cbind, lapply(seq_along(X_list), function(k) {
      s <- screens[[k]]
      X_block <- X_list[[k]][tr, s$genes, drop = FALSE]
      sweep(sweep(X_block, 2, s$center, "-"), 2, s$scale, "/")
    }))
    gene_block_te <- do.call(cbind, lapply(seq_along(X_list), function(k) {
      s <- screens[[k]]
      X_block <- X_list[[k]][i, s$genes, drop = FALSE]
      sweep(sweep(X_block, 2, s$center, "-"), 2, s$scale, "/")
    }))

    Xtr_full <- cbind(Age = age_tr, gene_block_tr)
    Xte_full <- cbind(Age = age_te, gene_block_te)
    if (nrow(Xte_full) != 1L) Xte_full <- matrix(Xte_full, nrow = 1L,
                                                  dimnames = list(NULL,
                                                                  colnames(Xtr_full)))

    pf <- c(0, rep(1, ncol(Xtr_full) - 1L))
    cvf <- tryCatch({
      cv.glmnet(Xtr_full, y_tr, family = "binomial",
                alpha = alpha_enet, penalty.factor = pf,
                nfolds = 5, type.measure = "deviance")
    }, error = function(e) {
      message("cv.glmnet failed at fold ", i, ": ", conditionMessage(e))
      NULL
    })
    if (is.null(cvf)) {
      p_hat[i] <- mean(y_tr)
    } else {
      p_hat[i] <- as.numeric(
        predict(cvf, newx = Xte_full, s = "lambda.min", type = "response")
      )
    }
  }
  list(p_hat = p_hat,
       n_kept = n_kept_log,
       fold_bal = fold_bal,
       screens = screens_log)
}
```

```{r run-loo, cache = FALSE}
y01 <- cohort$y01
age <- cohort$age

# (i) cfDNA only
res_cf  <- loo_predict(list(cf = X_cf), y01, age)
# (ii) gDNA only
res_g   <- loo_predict(list(g  = X_g),  y01, age)
# (iii) cfDNA + gDNA combined
res_cmb <- loo_predict(list(cf = X_cf, g = X_g), y01, age)

results <- list(
  `cfDNA only`         = res_cf,
  `gDNA only`          = res_g,
  `cfDNA + gDNA`       = res_cmb
)
```

\newpage

## Per-fold sanity checks

```{r fold-balance-table}
fb <- as.data.frame(res_cf$fold_bal) %>%
  mutate(Fold = seq_len(nrow(.)),
         `Total train` = MRDneg + MRDpos,
         `Held out` = ifelse(y01 == 0, "MRDneg", "MRDpos"))
knitr::kable(
  fb %>% select(Fold, MRDneg, MRDpos, `Total train`, `Held out`),
  format = "latex", booktabs = TRUE, escape = FALSE,
  caption = "Training-set class counts per LOO fold (same across all three models by construction)."
) %>% kable_styling(latex_options = c("HOLD_position"))
```

```{r fig5-fold-balance}
balance_df <- as.data.frame(res_cf$fold_bal) %>%
  mutate(Fold = seq_len(n())) %>%
  pivot_longer(cols = c(MRDneg, MRDpos),
               names_to = "Class", values_to = "Count")
p_balance <- ggplot(balance_df, aes(Fold, Count, fill = Class)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(MRDneg = "#457B9D", MRDpos = "#E63946")) +
  scale_x_continuous(breaks = 1:17) +
  labs(x = "LOO fold", y = "Patients in training set",
       title = "Per-fold training-class balance",
       subtitle = "Each fold leaves out one patient; the remaining 16 form the training set") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
save_both(p_balance, "fig_mrd_pred_fold_balance", width = 8, height = 4)
include_graphics(file.path(fig_out, "fig_mrd_pred_fold_balance.pdf"))
```

```{r fig4-gene-counts}
gene_count_df <- bind_rows(
  data.frame(Source = "cfDNA", n = res_cf$n_kept[, "cf"], Fold = 1:17),
  data.frame(Source = "gDNA",  n = res_g$n_kept[, "g"],   Fold = 1:17),
  data.frame(Source = "Combined: cfDNA",
             n = res_cmb$n_kept[, "cf"], Fold = 1:17),
  data.frame(Source = "Combined: gDNA",
             n = res_cmb$n_kept[, "g"],  Fold = 1:17)
)
gene_count_df$Source <- factor(gene_count_df$Source,
                               levels = c("cfDNA", "gDNA",
                                          "Combined: cfDNA",
                                          "Combined: gDNA"))
p_gene_counts <- ggplot(gene_count_df, aes(Source, n, fill = Source)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, height = 0, size = 1.5, alpha = 0.7) +
  scale_fill_manual(values = c("cfDNA" = "#E63946",
                               "gDNA"  = "#457B9D",
                               "Combined: cfDNA" = "#F4A6AB",
                               "Combined: gDNA"  = "#A8DADC"),
                    guide = "none") +
  labs(y = "Number of genes retained after screen",
       x = NULL,
       title = "Stage-2 (GMM top cluster) gene counts per LOO fold") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        panel.grid.minor = element_blank())
save_both(p_gene_counts, "fig_mrd_pred_gene_counts", width = 7, height = 4.5)
include_graphics(file.path(fig_out, "fig_mrd_pred_gene_counts.pdf"))
```

\newpage

## Inference and metrics

For each of the three models we compute:

- **AUROC** + DeLong 95% CI (`pROC::roc`, `pROC::ci.auc`).
- **Brier score** on the leave-one-out predicted probabilities.
- **Calibration intercept** via `glm(y ~ offset(logit(p_hat)))` and
  **calibration slope** via `glm(y ~ logit(p_hat))`, both with `family = binomial`.
- **Youden-optimal sensitivity, specificity, accuracy**
  (`pROC::coords(..., "best", best.method = "youden")`).
- **Pairwise DeLong tests** comparing AUROCs across the three models
  (`pROC::roc.test(..., method = "delong", paired = TRUE)`).

```{r metrics}
logit <- function(p) {
  p <- pmin(pmax(p, 1e-6), 1 - 1e-6)
  log(p / (1 - p))
}

compute_metrics <- function(p_hat, y01, model_name) {
  roc_obj <- pROC::roc(y01, p_hat, direction = "<", quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))
  ci_val  <- as.numeric(pROC::ci.auc(roc_obj, method = "delong"))
  brier   <- mean((p_hat - y01)^2)
  # Calibration intercept (CI): fix slope at 1, fit intercept
  fit_int <- tryCatch(
    glm(y01 ~ offset(logit(p_hat)), family = binomial()),
    error = function(e) NULL)
  ci_int  <- if (!is.null(fit_int)) {
    coef(summary(fit_int))[1, 1:2]
  } else c(NA, NA)
  # Calibration slope (CS): fit slope and intercept jointly
  fit_slp <- tryCatch(
    glm(y01 ~ logit(p_hat), family = binomial()),
    error = function(e) NULL)
  cs <- if (!is.null(fit_slp)) {
    coef(summary(fit_slp))[2, 1:2]
  } else c(NA, NA)
  # Youden-best cutoff
  cs_y <- tryCatch(
    pROC::coords(roc_obj, "best", best.method = "youden",
                 ret = c("sensitivity", "specificity", "accuracy", "threshold"),
                 transpose = FALSE),
    error = function(e) NULL)
  if (is.null(cs_y) || nrow(cs_y) == 0) {
    sens <- spec <- acc <- thr <- NA_real_
  } else {
    sens <- cs_y$sensitivity[1]
    spec <- cs_y$specificity[1]
    acc  <- cs_y$accuracy[1]
    thr  <- cs_y$threshold[1]
  }
  list(
    model      = model_name,
    roc        = roc_obj,
    auc        = auc_val,
    auc_ci_lo  = ci_val[1],
    auc_ci_hi  = ci_val[3],
    brier      = brier,
    calib_int  = ci_int[1],  calib_int_se = ci_int[2],
    calib_slp  = cs[1],      calib_slp_se = cs[2],
    sens       = sens, spec = spec, acc = acc, thr = thr
  )
}

metrics_list <- lapply(seq_along(results), function(k) {
  compute_metrics(results[[k]]$p_hat, y01, names(results)[k])
})
names(metrics_list) <- names(results)
```

```{r table2-headline}
fmt_ci <- function(est, lo, hi, digits = 3) {
  sprintf("%.3f (%.3f-%.3f)", est, lo, hi)
}
fmt_se <- function(est, se, digits = 3) {
  if (is.na(est)) return("--")
  sprintf("%.3f (%.3f)", est, se)
}

tab2 <- data.frame(
  Model        = names(metrics_list),
  `n genes (med, IQR)` = sapply(seq_along(results), function(k) {
    nks <- rowSums(results[[k]]$n_kept)
    sprintf("%.0f (%.0f-%.0f)",
            median(nks), quantile(nks, 0.25), quantile(nks, 0.75))
  }),
  `AUROC (95% CI)` = sapply(metrics_list, function(m)
    fmt_ci(m$auc, m$auc_ci_lo, m$auc_ci_hi)),
  Brier            = sapply(metrics_list, function(m) sprintf("%.3f", m$brier)),
  `Calib intercept (SE)` = sapply(metrics_list, function(m)
    fmt_se(m$calib_int, m$calib_int_se)),
  `Calib slope (SE)`     = sapply(metrics_list, function(m)
    fmt_se(m$calib_slp, m$calib_slp_se)),
  `Sens @ Youden` = sapply(metrics_list, function(m)
    sprintf("%.3f", m$sens)),
  `Spec @ Youden` = sapply(metrics_list, function(m)
    sprintf("%.3f", m$spec)),
  Accuracy = sapply(metrics_list, function(m) sprintf("%.3f", m$acc)),
  check.names = FALSE,
  stringsAsFactors = FALSE,
  row.names = NULL
)

knitr::kable(
  tab2, format = "latex", booktabs = TRUE, escape = TRUE,
  caption = "Headline LOO-CV performance metrics, by model. AUROC 95\\% CIs are DeLong intervals; calibration intercept is fit with slope fixed at 1; calibration slope is fit jointly with intercept. Sensitivity, specificity, and accuracy are evaluated at the Youden-optimal threshold."
) %>% kable_styling(latex_options = c("HOLD_position", "scale_down"))
```

```{r table3-delong}
roc_list <- lapply(metrics_list, function(m) m$roc)
pairs <- combn(length(roc_list), 2)
delong_p <- matrix(NA_real_, length(roc_list), length(roc_list),
                   dimnames = list(names(roc_list), names(roc_list)))
diag(delong_p) <- 1
for (k in seq_len(ncol(pairs))) {
  i <- pairs[1, k]; j <- pairs[2, k]
  test <- tryCatch(
    pROC::roc.test(roc_list[[i]], roc_list[[j]],
                   method = "delong", paired = TRUE),
    error = function(e) NULL
  )
  if (!is.null(test)) {
    delong_p[i, j] <- test$p.value
    delong_p[j, i] <- test$p.value
  }
}
delong_df <- as.data.frame(
  apply(delong_p, c(1, 2), function(x) {
    if (is.na(x)) "--" else if (x == 1) "--" else sprintf("%.3f", x)
  })
)
delong_df <- tibble::rownames_to_column(delong_df, "Model")
knitr::kable(
  delong_df, format = "latex", booktabs = TRUE, escape = TRUE,
  caption = "Pairwise DeLong test p-values for AUROC differences (paired across LOO predictions). Diagonal omitted."
) %>% kable_styling(latex_options = c("HOLD_position"))
```

\newpage

## Figures: ROC, AUROC forest, calibration

```{r fig1-roc}
roc_df <- bind_rows(lapply(seq_along(metrics_list), function(k) {
  r <- metrics_list[[k]]$roc
  data.frame(
    Model       = names(metrics_list)[k],
    Sensitivity = rev(r$sensitivities),
    Specificity = rev(r$specificities)
  )
}))
roc_df$FPR <- 1 - roc_df$Specificity
roc_df$Model <- factor(roc_df$Model, levels = names(metrics_list))

label_df <- data.frame(
  Model = names(metrics_list),
  label = sapply(metrics_list, function(m)
    sprintf("AUC = %.2f (%.2f-%.2f)", m$auc, m$auc_ci_lo, m$auc_ci_hi))
)

model_palette <- c("cfDNA only"   = "#E63946",
                   "gDNA only"    = "#457B9D",
                   "cfDNA + gDNA" = "#2C7A4B")

p_roc <- ggplot(roc_df, aes(FPR, Sensitivity, color = Model)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey60") +
  geom_path(linewidth = 0.9) +
  scale_color_manual(values = model_palette) +
  coord_equal() +
  labs(x = "False positive rate", y = "True positive rate",
       title = "Leave-one-out ROC curves",
       subtitle = "AUROCs with DeLong 95% CIs shown in legend") +
  theme_bw(base_size = 11) +
  theme(legend.position = c(0.65, 0.18),
        legend.background = element_rect(fill = "white", color = "grey60"),
        panel.grid.minor = element_blank()) +
  guides(color = guide_legend(
    title = NULL,
    override.aes = list(linewidth = 1.2),
    label.theme = element_text(size = 9)
  ))
# Add custom legend labels with AUC text
p_roc <- p_roc + scale_color_manual(
  values = model_palette,
  labels = setNames(label_df$label, label_df$Model)
)
save_both(p_roc, "fig_mrd_pred_roc", width = 6, height = 6)
include_graphics(file.path(fig_out, "fig_mrd_pred_roc.pdf"))
```

```{r fig2-forest}
forest_df <- data.frame(
  Model = factor(names(metrics_list), levels = rev(names(metrics_list))),
  AUC   = sapply(metrics_list, function(m) m$auc),
  lo    = sapply(metrics_list, function(m) m$auc_ci_lo),
  hi    = sapply(metrics_list, function(m) m$auc_ci_hi)
)
p_forest <- ggplot(forest_df, aes(AUC, Model, color = Model)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.18, linewidth = 0.7) +
  geom_point(size = 3.5) +
  scale_color_manual(values = model_palette, guide = "none") +
  xlim(0, 1) +
  labs(x = "AUROC (95% DeLong CI)", y = NULL,
       title = "AUROC forest plot",
       subtitle = "Vertical dashed line: AUROC = 0.5 (chance)") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
save_both(p_forest, "fig_mrd_pred_auc_forest", width = 7, height = 3.5)
include_graphics(file.path(fig_out, "fig_mrd_pred_auc_forest.pdf"))
```

```{r fig3-calibration}
calib_df <- bind_rows(lapply(seq_along(results), function(k) {
  data.frame(
    Model = names(results)[k],
    p_hat = results[[k]]$p_hat,
    y     = y01
  )
}))
calib_df$Model <- factor(calib_df$Model, levels = names(metrics_list))

p_calib <- ggplot(calib_df, aes(p_hat, y, color = Model)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey60") +
  geom_smooth(method = "loess", se = TRUE, span = 1.2,
              method.args = list(family = "gaussian"), linewidth = 0.7) +
  geom_point(size = 1.8, alpha = 0.7) +
  scale_color_manual(values = model_palette) +
  facet_wrap(~ Model, nrow = 1) +
  labs(x = "Predicted probability (LOO)",
       y = "Observed MRD-positive status",
       title = "Calibration of LOO predicted probabilities",
       subtitle = "Loess fit with 95% confidence band; dashed line = perfect calibration") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "grey92"),
        panel.grid.minor = element_blank())
save_both(p_calib, "fig_mrd_pred_calibration", width = 10, height = 4)
include_graphics(file.path(fig_out, "fig_mrd_pred_calibration.pdf"))
```

\newpage

## Supplementary: gene retention frequencies across LOO folds

The two-stage screen is re-run on each LOO training set; a gene that is retained
in many folds is more likely to be biologically informative than one retained
in a single fold by chance. We show below the genes retained in $\geq 50\%$ of
the 17 LOO folds, separately for cfDNA and gDNA (cfDNA-only and gDNA-only model
folds, respectively, to avoid the combined model's inflation).

```{r table-s1-cfdna-genes}
tally_genes <- function(screens_log, source_key = NULL) {
  all_genes <- unlist(lapply(screens_log, function(s) {
    if (is.null(source_key)) unlist(s) else s[[source_key]]
  }))
  freq <- table(all_genes)
  data.frame(
    gene = names(freq),
    n_folds = as.integer(freq),
    pct_folds = 100 * as.numeric(freq) / length(screens_log),
    stringsAsFactors = FALSE
  ) %>% arrange(desc(n_folds))
}

cf_tally <- tally_genes(res_cf$screens, "cf")
g_tally  <- tally_genes(res_g$screens,  "g")

cf_top <- cf_tally %>% filter(n_folds >= 9) %>% mutate(pct_folds = sprintf("%.0f%%", pct_folds))
g_top  <- g_tally  %>% filter(n_folds >= 9) %>% mutate(pct_folds = sprintf("%.0f%%", pct_folds))

if (nrow(cf_top) == 0) {
  cat("\\textit{No cfDNA genes retained in at least 9 of 17 LOO folds.}\n")
} else {
  knitr::kable(
    cf_top, format = "latex", booktabs = TRUE, escape = TRUE,
    col.names = c("Gene", "Folds retained (n)", "Folds retained (%)"),
    caption = "Table S1. cfDNA-only model: genes retained in $\\geq 9$ of 17 LOO folds."
  ) %>% kable_styling(latex_options = c("HOLD_position"))
}
```

```{r table-s2-gdna-genes}
if (nrow(g_top) == 0) {
  cat("\\textit{No gDNA genes retained in at least 9 of 17 LOO folds.}\n")
} else {
  knitr::kable(
    g_top, format = "latex", booktabs = TRUE, escape = TRUE,
    col.names = c("Gene", "Folds retained (n)", "Folds retained (%)"),
    caption = "Table S2. gDNA-only model: genes retained in $\\geq 9$ of 17 LOO folds."
  ) %>% kable_styling(latex_options = c("HOLD_position"))
}
```

\newpage

## Headline summary

```{r headline-text, results = "asis"}
cat("Per-model leave-one-out performance and pairwise DeLong tests:\n\n")
for (k in seq_along(metrics_list)) {
  m <- metrics_list[[k]]
  cat(sprintf("- **%s**: AUROC = %.3f (95%% DeLong CI %.3f to %.3f); Brier = %.3f; Sens@Youden = %.3f; Spec@Youden = %.3f.\n",
              names(metrics_list)[k], m$auc, m$auc_ci_lo, m$auc_ci_hi,
              m$brier, m$sens, m$spec))
}
cat("\nPairwise DeLong p-values:\n\n")
for (k in seq_len(ncol(pairs))) {
  i <- pairs[1, k]; j <- pairs[2, k]
  cat(sprintf("- **%s vs %s**: p = %s.\n",
              names(metrics_list)[i], names(metrics_list)[j],
              if (is.na(delong_p[i, j])) "NA" else sprintf("%.3f", delong_p[i, j])))
}
```

## Caveats and interpretation

- **Sample size**: $n = 17$ with a 7-vs-10 MRD class split is severely
  underpowered. Expected DeLong 95% CI width on a single AUROC is ~0.30; the
  CI on an AUROC *difference* between two models is even wider. Non-significant
  pairwise DeLong tests are consistent with both genuine equivalence and an
  underpowered comparison; both readings are admissible.

- **Stage 1 multiple testing**: with ~11,000 genes per source and a Stage-1
  threshold of uncorrected $p < 0.05$, we expect ~550 false positives per fold.
  The Stage-2 GMM step prunes most of these but is *not* formal error control.
  The genes-retained-in-many-folds tables (S1, S2) are the closest thing to a
  stable signal in this analysis.


- **Calibration is fragile at this $n$**: a calibration slope CI can easily span
  zero with 17 patients. Calibration intercept/slope are reported for
  completeness but should not drive decisions.


# Note on the removed per-component L1 analysis

Sections that regressed the per-component $L_1$ concordance $C_{it}^{(k)}$ on
the MRD trajectory (stacked functional regression and per-cycle ANOVA) have
been removed. The $L_1$ concordance metric is no longer reported in this
paper; the single subclonal-concordance measure is the posterior correlation
of $\omega^{cf}$ and $\omega^{g}$ (see `R/04_concordance_metrics.R`,
`krd_posterior_correlation.csv`), and its association with MRD is reported in
`R/11_concordance_mrd_association.R`.


# Session info

```{r sessioninfo, echo = TRUE}
sessionInfo()
```
